sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-18 doc/paper.Rmd
Develop example

We must then create a mesh object that contains matrices to apply the SPDE approach:

@@

Here, `cutoff` defines the minimum allowed distance between points.
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh(mesh = ...)`. 
We can then inspect our mesh object with the associated `plot()` method: `plot(mesh)` (see Fig. \@ref(fig:pcod-fig)a):

```{r, eval=FALSE, echo=FALSE}
plot(mesh)
```

And use our data frame and mesh to fit a spatial model:

```{r pcod-eg1-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE}
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod,
  family = tweedie(link = "log"),
  mesh = mesh,
  spatial = 'on'
)
```

@@

The output indicates our model was fit by (marginal) maximum likelihood ('ML').
We also see the formula, mesh, fitted data, and family.
Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the \mbox{Mat\'ern} range distance, marginal spatial standard deviation, and the negative log likelihood at convergence.

We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
The data frame `qcs_grid` is a grid that extends over the full studied domain and contains all predictors used in the model set at values we wish to predict at.

@@

We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.

@@

```{r}
g_pred <- ggplot(p, aes(X, Y, fill = exp(est))) + geom_raster() +
  scale_fill_viridis_c(trans = "sqrt", option = "C") + coord_fixed() +
  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
```

```{r, eval=FALSE}
p3 <- predict(fit, newdata = subset(qcs_grid, year == 2003), sims = 500L)
se <- apply(p3, 1, sd)
p$se <- se
g_pred_se <- ggplot(p, aes(X, Y, fill = se)) + geom_raster() +
  scale_fill_viridis_c(trans = "log10", option = "C") + coord_fixed() +
  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))
```

@@

```{r}
nd <- data.frame(depth =
  seq(min(pcod$depth), max(pcod$depth), length.out = 300))
p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
g_smooth <- ggplot(p_smooth,
  aes(depth, exp(est), ymin = exp(est - 1.96 * est_se),
    ymax = exp(est + 1.96 * est_se))) +
  geom_ribbon(fill = blues[2]) +
  geom_line(colour = blues[5], lwd = 1) +
  labs(x = "Depth (m)", y = "Density (units)") +
  xlim(min(pcod$depth), 300) +
  coord_fixed(expand = FALSE, ratio = 1.28)
```

```{r, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=TRUE}
fit2 <- sdmTMB(
  density ~ s(depth), 
  family = tweedie(link = "log"), data = pcod, mesh = mesh,
  time = 'year', spatial = 'off', spatiotemporal = 'ar1'
)
```

```{r fit-pcod-st, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=FALSE}
fit2 <- sdmTMB(
  density ~ 0 + as.factor(year), 
  family = tweedie(link = "log"), data = sdmTMB::pcod, mesh = mesh,
  time = 'year', spatial = 'on', spatiotemporal = 'iid', silent = TRUE
)
p2 <- predict(fit2, newdata = qcs_grid, sims = 500L)
ind <- get_index_sims(p2)
```

@@

```{r, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=TRUE}
fit_spatiotemporal <- sdmTMB(
  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh
  time = 'year', spatial = 'off', spatiotemporal = 'ar1'
)
```

>>>>

We start by creating a mesh object that contains matrices to apply the SPDE approach:

@@

Here, `cutoff` defines the minimum allowed distance between points, here in km.
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh(mesh = ...)`. 
We can then inspect our mesh object with the associated `plot()` method: `plot(mesh)` (see Fig. \@ref(fig:pcod-fig)a).

We can then fit a model with spatial random fields via the function `sdmTMB()`. 
We use a penalized smoother for depth as a main effect via `mgcv::s()`.
We specify the family as Tweedie to account for the positive continuous density values that also have zeros.

```{r pcod-eg1-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE}
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod,
  family = tweedie(link = "log"),
  mesh = mesh,
  spatial = "on"
)
```

@@

The output indicates our model was fit by (marginal) maximum likelihood (`'ML'`).
We also see the formula, mesh, fitted data, and family.
Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the \mbox{Mat\'ern} range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.

We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
The data frame `qcs_grid` is a grid that extends over the full studied domain and contains all predictors used in the model set at values we wish to predict at.
We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.

@@

@@

```{r}
g_pred <- ggplot(p, aes(X, Y, fill = exp(est))) +
  geom_raster() +
  scale_fill_viridis_c(trans = "sqrt", option = "C") +
  coord_fixed() +
  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
```

```{r, eval=FALSE}
p3 <- predict(fit, newdata = subset(qcs_grid, year == 2003), sims = 500L)
se <- apply(p3, 1, sd)
p$se <- se
g_pred_se <- ggplot(p, aes(X, Y, fill = se)) +
  geom_raster() +
  scale_fill_viridis_c(trans = "log10", option = "C") +
  coord_fixed() +
  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
```

@@

```{r}
nd <- data.frame(
  depth =
    seq(min(pcod$depth), max(pcod$depth), length.out = 300)
)
p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
g_smooth <- ggplot(
  p_smooth,
  aes(depth, exp(est),
    ymin = exp(est - 1.96 * est_se),
    ymax = exp(est + 1.96 * est_se)
  )
) +
  geom_ribbon(fill = blues[2]) +
  geom_line(colour = blues[5], lwd = 1) +
  labs(x = "Depth (m)", y = "Density (units)") +
  xlim(min(pcod$depth), 300) +
  coord_fixed(expand = FALSE, ratio = 1.28)
```

```{r fit-pcod-st, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=FALSE}
fit2 <- sdmTMB(
  density ~ 0 + as.factor(year),
  family = tweedie(link = "log"), data = sdmTMB::pcod, mesh = mesh,
  time = "year", spatial = "on", spatiotemporal = "iid", silent = TRUE
)
p2 <- predict(fit2, newdata = qcs_grid, sims = 500L)
ind <- get_index_sims(p2)
```

@@

```{r, eval=FALSE, echo=TRUE}
fit_spatiotemporal <- sdmTMB(
  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh
  time = "year", spatial = "off", spatiotemporal = "ar1"
)
```
